---
title: "Evaluating epigenetic clocks and scores: data preparation"
author: "Vera N. Karlbauer"
contact: "vera_karlbauer@psych.mpg.de"
date: "2024-12-03"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/vera_karlbauer/Documents/Kids2Health/02_stats/02_analysis/evaluating_epigenetic_clocks_and_scores")
```

# Purpose:
- Aggregate and prepare epigenetic clocks and scores from blood and saliva in Kids2Health cohorts SP1/2/4. 
- Create versions of all epigenetic clocks, scores, and acceleration measures residualized against the respective estimated cell type proportions (epithelial cells for saliva, everything but neutrophils for blood). 
- Export complete dataset and "inner join" (children with both tissues only). 

## Step 0: Load packages

```{r}
# load packages
library(rmarkdown)
library(dplyr)
library(tibble)
library(janitor)
library(readxl)
library(ggplot2)
```

## Step 1: Load data

```{r}
## epigenetic scores
# blood
load("../../../01_data/02_proc/04_epi_scores/blood/epigenetic_bmi_score_do_2023_K2H_SP124_blood.Rdata")
load("../../../01_data/02_proc/04_epi_scores/blood/epigenetic_bmi_score_wahl_2017_K2H_SP124_blood.Rdata")
load("../../../01_data/02_proc/04_epi_scores/blood/epigenetic_crp_score_wielscher_2022_K2H_SP124_blood.Rdata")
load("../../../01_data/02_proc/04_epi_scores/blood/epigenetic_crp_score_ligthart_2016_K2H_SP124_blood.Rdata")
load("../../../01_data/02_proc/04_epi_scores/blood/epigenetic_g_K2H_SP124_blood.Rdata")
load("../../../01_data/02_proc/04_epi_scores/blood/smoking_exposure_K2H_SP124_blood.Rdata")
# saliva
load("../../../01_data/02_proc/04_epi_scores/saliva/epigenetic_bmi_score_do_2023_K2H_SP124_saliva.Rdata")
load("../../../01_data/02_proc/04_epi_scores/saliva/epigenetic_bmi_score_wahl_2017_K2H_SP124_saliva.Rdata")
load("../../../01_data/02_proc/04_epi_scores/saliva/epigenetic_crp_score_wielscher_2022_K2H_SP124_saliva.Rdata")
load("../../../01_data/02_proc/04_epi_scores/saliva/epigenetic_crp_score_ligthart_2016_K2H_SP124_saliva.Rdata")
load("../../../01_data/02_proc/04_epi_scores/saliva/epigenetic_g_K2H_SP124_saliva.Rdata")
load("../../../01_data/02_proc/04_epi_scores/saliva/smoking_exposure_K2H_SP124_saliva.Rdata")

## epigenetic clocks
# blood
load("../../../01_data/02_proc/04_epi_scores/blood/epigenetic_clocks_K2H_SP124_blood.Rdata")
# saliva
load("../../../01_data/02_proc/04_epi_scores/saliva/epigenetic_clocks_K2H_SP124_saliva.Rdata")

## cell types
# saliva
load("../../../01_data/02_proc/01_epic/kids2health_sp1_2_4_saliva/processed_data/final_data/celltypes_unfiltered_saliva_MiddletonPanel.Rdata")
celltypes_unfiltered_saliva <- celltypes_unfiltered
# blood
load("../../../01_data/02_proc/01_epic/kids2health_sp1_2_4_blood_new/processed_data/final_data/celltypes_unfiltered.Rdata")

# phenotypes
load("../../../01_data/02_proc/05_complete_phenotypes/full_phenotypes_K2H_SP124.Rdata")

# Nightingale Plasma Metabolomics Panel (for total cholesterol, HDL-C, LDL-C, total triglycerides)
metabolomics <- read_excel("../../../01_data//01_raw/SP_1_2_4_hormones_inflammation_nightingale/K2H nightingale panel.xlsx") 

# blood betas (restrict to smoking CpG immediately)
load("../../../01_data/02_proc/01_epic/kids2health_sp1_2_4_blood_new/processed_data/final_data/Betas_clean_unfiltered_quantile_bmiq_combated_pseudo_v1_final.Rdata") 
smoking_cpg_blood <- as.data.frame(t(Betas_clean_unfiltered_quantile_bmiq_combated_pseudo_v1)) %>%
  rownames_to_column(var = "arrayid") %>%
  select(arrayid, cg05575921)
remove(Betas_clean_unfiltered_quantile_bmiq_combated_pseudo_v1)
  
# saliva betas (restrict to smoking CpG immediately)
load("../../../01_data/02_proc/01_epic/kids2health_sp1_2_4_saliva/processed_data/final_data/Betas_clean_unfiltered_quantile_bmiq_combated_pseudo_v1_final.Rdata")
smoking_cpg_saliva <- as.data.frame(t(Betas_clean_unfiltered_quantile_bmiq_combated_pseudo_v1)) %>%
  rownames_to_column(var = "arrayid") %>%
  select(arrayid, cg05575921)
remove(Betas_clean_unfiltered_quantile_bmiq_combated_pseudo_v1)
```

## Step 2: Process phenotype data

```{r}
## filter phenotypes and create surrogate variables
full_phenotypes <- full_phenotypes %>%
  select(child_id, zeitpunkt, group, ses_total,
         corona_ksl_01, corona_ksl_mix01, fasd_cat, ku_bmi, bmi_percentile, 
         iq_sonr_nonverbal, iq_wisc, crp_pgML, telomere_length_blood_ts_ratio,
         ku_blutdr1_sys, ku_blutdr2_sys, ku_blutdr_sys_mw, 
         ku_blutdr_dias_mw, ku_blutdr1_dias, ku_blutdr2_dias,
         bodpodfm_i, bioproben_blood4) %>%
  # make sure trauma group is not-numeric
  mutate(group = as.factor(group)) %>%
  # recode & rename ethnicity variables
  rename(ethnicity = corona_ksl_01,
         ethnicity_mixed_spec = corona_ksl_mix01) %>%
  mutate(ethnicity = case_match(ethnicity, 1 ~ "White", 2 ~ "Black", 3 ~ "Asian", 4 ~ "Mixed", 5 ~ "Other"),
         ethnicity_mixed_spec = case_match(ethnicity_mixed_spec, 1 ~ "Mixed: White/Black", 2 ~ "Mixed: Asian/Black",
                                           3 ~ "Mixed: White/Asian", 4 ~ "Mixed: Other")) %>%
  # combine ethnicity variables into one variable
  mutate(ethnicity = case_when(ethnicity == "Mixed" ~ ethnicity_mixed_spec, TRUE ~ ethnicity)) %>%
  select(-ethnicity_mixed_spec) %>%
  # extract fasting status at blood sample from free text comments
  mutate(blood_sample_fasted = case_when(grepl("nicht nüchtern", bioproben_blood4) == TRUE ~ FALSE,
                                          TRUE ~ TRUE)) %>%
  # log-normalize CRP
  mutate(crp_pgML_ln = log(crp_pgML)) %>%
  # rename health biomarkers
  rename(bp_systolic = ku_blutdr_sys_mw,
         bp_diastolic = ku_blutdr_dias_mw,
         bodpod_fatmass = bodpodfm_i)

## blood pressure QC
# flag systolic < diastolic
full_phenotypes %>%
  filter(bp_systolic < bp_diastolic) %>%
  nrow()
full_phenotypes %>%
  filter(ku_blutdr1_sys < ku_blutdr1_dias) %>%
  nrow()
full_phenotypes %>%
  filter(ku_blutdr2_sys < ku_blutdr2_dias) %>%
  nrow()
# flag biologically implausible averages
full_phenotypes %>%
  filter(bp_systolic >200) %>%
  nrow()
full_phenotypes %>%
  filter(bp_systolic <70) %>%
  nrow()
# flag range between 2 repeated measurements exceeding >30
full_phenotypes %>%
  filter(abs(ku_blutdr1_sys - ku_blutdr2_sys) >30) %>%
  mutate(range_systolic = abs(ku_blutdr1_sys - ku_blutdr2_sys)) %>%
  nrow()
# exclude observation(s) with range between 2 repeated measurements exceeding >30
full_phenotypes <- full_phenotypes %>%
  mutate(bp_systolic = case_when(abs(ku_blutdr1_sys - ku_blutdr2_sys) >30 ~ NA,
                                 TRUE ~ bp_systolic))
# remove unnecessary bp variables
full_phenotypes <- full_phenotypes %>%
  select(-c(ku_blutdr1_sys, ku_blutdr2_sys, ku_blutdr1_dias, ku_blutdr2_dias, bp_diastolic))
```

### Step 3: Process metabolomics

```{r}
## process metabolomics data
metabolomics <- metabolomics %>%
  # recode ID, set timepoint to T0 (not sampled at T2)
  mutate(child_id = tolower(kids2health_id),
         timepoint = rep("T0", length(child_id))) %>%
  # extract total cholesterol, HDL-C, LDL-C, total triglycerides
  select(child_id, timepoint, TotalC, LDLC, HDLC, TotalTG) %>%
  # remove NAs
  filter(!(is.na(TotalC) & is.na(LDLC) & is.na(HDLC) & is.na(TotalTG))) %>%
  # remove children who did not fast before blood sample (N=5)
  filter(!(child_id %in% filter(full_phenotypes, zeitpunkt == "T0" 
                                & blood_sample_fasted == FALSE)$child_id))

## detect metabolite outliers >5SD from mean
# visualize outliers
metabolites <- c("TotalC", "LDLC", "HDLC", "TotalTG")
for(i in 1:4) {
  current_metabolite = metabolites[i]
  current_data = rename(metabolomics, "metabolite" = current_metabolite)
  plot = ggplot(data = current_data,
       aes(y = metabolite)) +
    geom_boxplot() +
    geom_hline(yintercept = c((mean(current_data$metabolite,
                                 na.rm = TRUE) + 
                              5*sd(current_data$metabolite,
                                   na.rm = TRUE)),
                           (mean(current_data$metabolite,
                                 na.rm = TRUE) - 
                              5*sd(current_data$metabolite,
                                   na.rm = TRUE))),
             color = "red", linetype = "dashed") +
    ylim(c(min(current_data$metabolite), max(current_data$metabolite))) +
    ggtitle(current_metabolite) +
    theme_bw()
  print(plot)
}
```

## Step 4: Aggregate blood data

```{r}
all_blood <- full_join(k2h_blood_epiclocks, epigenetic_bmi_score_do_2023_blood, by = c("sample_ID", "arrayid", "child_id", "zeitpunkt"))
all_blood <- full_join(all_blood, epigenetic_bmi_score_wahl_2017_blood, by = c("sample_ID", "arrayid", "child_id", "zeitpunkt"))
all_blood <- full_join(all_blood, epigenetic_crp_score_wielscher_2022_blood, by = c("sample_ID", "arrayid", "child_id", "zeitpunkt"))
all_blood <- full_join(all_blood, epigenetic_crp_score_ligthart_2016_blood, by = c("sample_ID", "arrayid", "child_id", "zeitpunkt"))
all_blood <- full_join(all_blood, epigenetic_g, by = c("sample_ID", "arrayid"))
all_blood <- full_join(all_blood, smoking_exposure, by = c("sample_ID", "arrayid"))
all_blood <- full_join(all_blood, smoking_cpg_blood, by = c("arrayid"))
all_blood <- full_join(all_blood, celltypes_unfiltered, by = c("sample_ID", "arrayid", "child_id", "zeitpunkt"  = "timepoint", "teilprojekt" = "subproject"))
```

## Step 3: Aggregate saliva data

```{r}
all_saliva <- full_join(k2h_saliva_epiclocks, epigenetic_bmi_score_do_2023_saliva, by = c("sample_ID", "arrayid", "child_id", "zeitpunkt"))
all_saliva <- full_join(all_saliva, epigenetic_bmi_score_wahl_2017_saliva, by = c("sample_ID", "arrayid", "child_id", "zeitpunkt"))
all_saliva <- full_join(all_saliva, epigenetic_crp_score_wielscher_2022_saliva, by = c("sample_ID", "arrayid", "child_id", "zeitpunkt"))
all_saliva <- full_join(all_saliva, epigenetic_crp_score_ligthart_2016_saliva, by = c("sample_ID", "arrayid", "child_id", "zeitpunkt"))
all_saliva <- full_join(all_saliva, epigenetic_g_saliva, by = c("sample_ID", "arrayid", "child_id", "zeitpunkt"))
all_saliva <- full_join(all_saliva, smoking_exposure_score_saliva, by = c("sample_ID", "arrayid", "child_id", "zeitpunkt"))
all_saliva <- full_join(all_saliva, smoking_cpg_saliva, by = c("arrayid"))
all_saliva <- full_join(all_saliva, celltypes_unfiltered_saliva, by = c("arrayid"))
```

## Step 5: Create cell type residualized versions of scores/clocks

```{r}
## blood
# prepare data
blood4correction <- all_blood %>%
  select(sample_ID, 
         starts_with("epiage"),
         starts_with("epigenetic"),
         ends_with("score"),
         dunedin_pace,
         CD4Tnv, Baso, CD4Tmem, Bmem, Bnv, Treg, CD8Tmem, CD8Tnv, Eos, NK, Mono) %>%
  column_to_rownames(var = "sample_ID")
# residualize for all cell types except neutrophils
blood_celltype_corrected <- c()
for (i in 1:(ncol(blood4correction)-11)) {
  blood_temp <- lm(blood4correction[,i] ~ 
                     blood4correction$CD4Tnv + blood4correction$Baso + blood4correction$CD4Tmem + blood4correction$Bmem + blood4correction$Bnv + blood4correction$Treg + blood4correction$CD8Tmem + blood4correction$CD8Tnv + blood4correction$Eos + blood4correction$NK + blood4correction$Mono
                   )
  blood_celltype_corrected<-cbind(blood_celltype_corrected,blood_temp$residuals)
  }
# clean up and append
colnames(blood_celltype_corrected) <- paste0(colnames(blood4correction)[1:(ncol(blood4correction)-11)], "_resid")
rownames(blood_celltype_corrected) <- rownames(blood4correction)
blood_celltype_corrected <- as.data.frame(blood_celltype_corrected) %>%
  rownames_to_column(var = "sample_ID")
all_blood <- left_join(all_blood, blood_celltype_corrected)

## saliva
# prepare data
saliva4correction <- all_saliva %>%
  select(sample_ID, starts_with("epiage"),
         starts_with("epigenetic"),
         ends_with("score"),
         dunedin_pace,
         Epithelial) %>%
  column_to_rownames(var = "sample_ID")
# residualize for epithelial cells
clocks_saliva_celltype_corrected <- c()
for (i in 1:(ncol(saliva4correction)-1)) {
    formula = paste(colnames(saliva4correction)[i], "~ Epithelial")
    saliva_temp <- lm(data = saliva4correction, formula = formula, na.action = "na.exclude")
  clocks_saliva_celltype_corrected <- cbind(clocks_saliva_celltype_corrected,resid(saliva_temp))
}
# clean up and append
colnames(clocks_saliva_celltype_corrected) <- paste0(colnames(saliva4correction)[1:(ncol(saliva4correction)-1)], "_resid")
rownames(clocks_saliva_celltype_corrected) <- rownames(saliva4correction)
clocks_saliva_celltype_corrected <- as.data.frame(clocks_saliva_celltype_corrected) %>%
  rownames_to_column(var = "sample_ID")
all_saliva <- left_join(all_saliva, clocks_saliva_celltype_corrected)

```

## Step 6: Combine blood and saliva data & append phenotypes

```{r}
# combine tissues
all_cross_tissue <- full_join(all_blood, all_saliva, 
                              by = c("child_id", "zeitpunkt", "teilprojekt", "sex"),
                              suffix = c("_blood", "_saliva"))

# rename cell types
all_cross_tissue <- all_cross_tissue %>%
  rename(
    "CD4Tnv_blood" = "CD4Tnv",
         "Baso_blood" = "Baso",
         "CD4Tmem_blood" = "CD4Tmem",
         "Bmem_blood" = "Bmem",
         "Bnv_blood" = "Bnv",
         "Treg_blood" = "Treg",
         "CD8Tmem_blood" = "CD8Tmem",
         "CD8Tnv_blood" = "CD8Tnv",
         "Eos_blood" = "Eos",
         "NK_blood" = "NK",
         "Neu_blood" = "Neu",
         "Mono_blood" = "Mono",
         "Epithelial_saliva" = "Epithelial",
         "Leukocytes_saliva" = "Leukocytes")

# rename age, timepoint, subproject variables
all_cross_tissue <- all_cross_tissue %>%
  rename("timepoint" = "zeitpunkt",
         "subproject" = "teilprojekt")

# append phenotypes
all_cross_tissue <- left_join(all_cross_tissue, full_phenotypes, by = c("child_id", "timepoint" = "zeitpunkt"))

# append Nightingale metabolites
all_cross_tissue <- left_join(all_cross_tissue, metabolomics, by = c("child_id", "timepoint"))

# create chronological age variable (age at blood or saliva sample, only used for plotting)
# age at blood and saliva sampling only deviates for 3 observations (9-35 days difference) - chronological age is set to saliva age for plotting, but is not used for statistical analysis
all_cross_tissue <- all_cross_tissue %>%
  mutate(chrono_age = case_when(
    is.na(bio_blood_age) ~ bio_saliva_age,
    is.na(bio_saliva_age) ~ bio_blood_age, 
    TRUE ~ bio_saliva_age))
  
# split off children with both tissues only
all_cross_tissue_inner <- all_cross_tissue %>%
  filter(!is.na(epiage_horvath_blood) & !is.na(epiage_horvath_saliva))
```

## Step 7: Export data

```{r}
# full dataset
save(all_cross_tissue, file = "../../../01_data/02_proc/09_cross_tissue/all_cross_tissue.Rdata")

# children with both tissues only
save(all_cross_tissue_inner, file = "../../../01_data/02_proc/09_cross_tissue/all_cross_tissue_inner.Rdata")

```

