---
title: "Evaluating epigenetic clocks and scores: Exposure effects"
author: "Vera N. Karlbauer"
contact: "vera_karlbauer@psych.mpg.de"
date: "2025-07-16"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/vera_karlbauer/Documents/Kids2Health/02_stats/02_analysis/evaluating_epigenetic_clocks_and_scores", digits = 4, scipen = 6)
```

# Purpose:  
- Attempt to replicate commonly found effects of early-life exposures on DNAm measures in blood and saliva.  
- Compare effect sizes between tissues.
- Exposures: early-life adversity (ELA) and socioeconomic status (SES). 
- Epigenetic measures: 14 epigenetic clocks, 7 epigenetic scores.  
- Covariates: age, sex, cell types, smoking (cg05575921), kinship matrix (ancestry & relatedness).
- Plot results as barplot with cell type & smoking adjustment (Figure 6) and without cell type & smoking (Supplementary Figure 4).
- Export results as Supplementary Table 10.

# Setup
### Step 0: Load packages

```{r}
library(rmarkdown)
library(dplyr)
library(RColorBrewer)
library(coxme)
library(writexl)
library(tidyr)
library(ggplot2)
library(stringr)
library(ggpubr)
library(ggpattern)
```

### Step 1: Source functions and utilities

```{r}
source("./00_functions/01_functions.R")
source("./00_functions/02_utilities.R")
```

### Step 2: Load data

```{r, include = FALSE}
# main data
load("../../../01_data/02_proc/09_cross_tissue/all_cross_tissue_inner.Rdata")
# kinship matrix
load("../../../01_data/02_proc/02_genotypes/qc-imputation_output/genotype_data/01_qc/13_ancestry_rels/K2H_SP1_2_3_4_genetic_kinship_matrix.Rdata")
```

### Step 3: Prepare data 

```{r}
# flip SES (*-1) for ELS & SES effects in same direction
all_cross_tissue_inner <- all_cross_tissue_inner %>%
  mutate(ses_total_negative = ses_total * -1)
# exclude all children not in kinship matrix (no genotype data available = 2 kids = 4 observations), only keep T0 (baseline)
all_cross_tissue_inner <- all_cross_tissue_inner %>%
  filter(child_id %in% rownames(k2H_pcrelate_kinship_matrix)) %>%
  filter(timepoint == "T0")

# filter kinship matrix
kinship <- k2H_pcrelate_kinship_matrix[unique(all_cross_tissue_inner$child_id), unique(all_cross_tissue_inner$child_id)]
```

### Create list of epigenetic measures of interest: EpiAgeAccel measures, DunedinPACE, epigenetic BMI & CRP, epigenetic-g, DNAmTL, PC DNAmTL.

```{r}
epimeasures <- gsub("_saliva", "", 
                    colnames(all_cross_tissue_inner %>%
                               select(contains("saliva")) %>%
                               select(starts_with("epiage_accel") 
                                      | contains("dunedin_pace")
                                      | contains("bmi") 
                                      | contains("crp") 
                                      | contains("epigenetic_g")
                                      | contains("epiage_telo")
                                      | contains("epiage_pc_telo")) %>%
                               select(!contains("resid"))))
```


# Analysis
## Early-life adversity (ELA) effects 
### Fit linear models in blood & saliva with and without correction for estimated cell type proportions and smoking exposure

```{r}
# create empty results object
results_ela <- data.frame(matrix(nrow = 0, ncol = 9))
colnames(results_ela) <- c("epigenetic_measure", 
                           "beta_std_blood_minimal_model", "p_blood_minimal_model", 
                           "beta_std_blood_smoking_celltype_adjusted",
                           "p_blood_smoking_celltype_adjusted", 
                           "beta_std_saliva_minimal_model", "p_saliva_minimal_model",
                           "beta_std_saliva_smoking_celltype_adjusted",
                           "p_saliva_smoking_celltype_adjusted")

# loop linear model over all epigenetic measures in blood and saliva
for(epimeasure in epimeasures){
  # subset blood data
  current_data_blood = select(all_cross_tissue_inner, 
                              paste0(epimeasure, "_blood"), child_id, group,
                              bio_blood_age, sex, 
                              cg05575921_blood,
                              CD4Tnv_blood, Baso_blood, CD4Tmem_blood, Bmem_blood, 
                              Bnv_blood, Treg_blood, CD8Tmem_blood, CD8Tnv_blood, 
                              Eos_blood, NK_blood, Mono_blood) %>%
    dplyr::rename("epimeasure" = paste0(epimeasure, "_blood")) %>%
    # standardize all numerical variables to mean = 0, sd = 1 (to get standardized betas)
    mutate(across(where(is.numeric), scale))
  # fit basic model in blood (only adjusted for age, sex, kinship)
  fit_blood_minimal = lmekin(
    formula = epimeasure ~ group + bio_blood_age + sex +
         (1 | child_id),
       data = current_data_blood,
       na.action = "na.omit", 
       varlist = kinship)
  sumstats_blood_minimal = extract_coxme_table(fit_blood_minimal)
  # extract results blood minimal model
  coefs = data.frame(matrix(nrow = 1, ncol = 0))
  coefs$epigenetic_measure = epimeasure
  coefs$beta_std_blood_minimal_model = sumstats_blood_minimal["group",]$beta
  coefs$p_blood_minimal_model = sumstats_blood_minimal["group",]$p
  # fit model in blood adjusted for age, sex, kinship, epigenetic smoke exposure proxy, estimated cell types
  fit_blood_full = lmekin(
    formula = epimeasure ~ group + bio_blood_age + sex +
      cg05575921_blood +
    CD4Tnv_blood + Baso_blood + CD4Tmem_blood + Bmem_blood + Bnv_blood + Treg_blood + 
    CD8Tmem_blood + CD8Tnv_blood + Eos_blood + NK_blood + Mono_blood +
         (1 | child_id),
       data = current_data_blood,
       na.action = "na.omit", 
       varlist = kinship)
  sumstats_blood_full = extract_coxme_table(fit_blood_full)
  # extract results blood full model
  coefs$epigenetic_measure = epimeasure
  coefs$beta_std_blood_smoking_celltype_adjusted = sumstats_blood_full["group",]$beta
  coefs$p_blood_smoking_celltype_adjusted = sumstats_blood_full["group",]$p
  # subset saliva data
  current_data_saliva = select(all_cross_tissue_inner, 
                        paste0(epimeasure, "_saliva"), child_id, group,
                        bio_saliva_age, sex, 
                        cg05575921_saliva, smoking_exposure_score_saliva,
                        Epithelial_saliva) %>%
    dplyr::rename("epimeasure" = paste0(epimeasure, "_saliva")) %>%
    # standardize all numerical variables to mean = 0, sd = 1 (to get standardized betas)
    mutate(across(where(is.numeric), scale))
  # fit basic model in saliva (only adjusted for age, sex, kinship)
  fit_saliva_minimal = lmekin(
    formula = epimeasure ~ group + bio_saliva_age + sex +
         (1 | child_id),
       data = current_data_saliva,
       na.action = "na.omit", 
       varlist = kinship)
  sumstats_saliva_minimal = extract_coxme_table(fit_saliva_minimal)
  # extract results saliva minimal model
  coefs$epigenetic_measure = epimeasure
  coefs$beta_std_saliva_minimal_model = sumstats_saliva_minimal["group",]$beta
  coefs$p_saliva_minimal_model = sumstats_saliva_minimal["group",]$p
  # fit model in saliva adjusted for age, sex, kinship, epigenetic smoke exposure proxy, estimated cell types
  fit_saliva_full = lmekin(
    formula = epimeasure ~ group + bio_saliva_age + sex +
      cg05575921_saliva +
      Epithelial_saliva +
         (1 | child_id),
       data = current_data_saliva,
       na.action = "na.omit", 
       varlist = kinship)
  sumstats_saliva_full = extract_coxme_table(fit_saliva_full)
  # extract results saliva full model
  coefs$epigenetic_measure = epimeasure
  coefs$beta_std_saliva_smoking_celltype_adjusted = sumstats_saliva_full["group",]$beta
  coefs$p_saliva_smoking_celltype_adjusted = sumstats_saliva_full["group",]$p
  # append results
  results_ela <- rbind(results_ela, coefs)
}

# FDR-correct results
fdr_ela <- as.data.frame(matrix(nrow = 2*nrow(results_ela), ncol = 0))
fdr_ela$tissue <- c(rep("blood", nrow(results_ela)), 
                   rep("saliva", nrow(results_ela)))
fdr_ela$p_fdr_minimal_model <- p.adjust(c(results_ela$p_blood_minimal_model, 
                                          results_ela$p_saliva_minimal_model), method = "fdr")
fdr_ela$p_fdr_smoking_celltype_adjusted <- p.adjust(c(results_ela$p_blood_smoking_celltype_adjusted,                                                    results_ela$p_saliva_smoking_celltype_adjusted), method = "fdr")
# append FDR-corrected results
results_ela <- results_ela %>%
  mutate(p_fdr_blood_minimal_model = filter(fdr_ela, tissue == "blood")$p_fdr_minimal_model,
         p_fdr_saliva_minimal_model = filter(fdr_ela, tissue == "saliva")$p_fdr_minimal_model,
         p_fdr_blood_smoking_celltype_adjusted = filter(fdr_ela, tissue == "blood")$p_fdr_smoking_celltype_adjusted,
         p_fdr_saliva_smoking_celltype_adjusted = filter(fdr_ela, tissue == "saliva")$p_fdr_smoking_celltype_adjusted) %>%
  relocate(p_fdr_blood_minimal_model, .after = "p_blood_minimal_model") %>%
  relocate(p_fdr_blood_smoking_celltype_adjusted, .after = "p_blood_smoking_celltype_adjusted")

# rename epigenetic clocks/scores
results_ela <- results_ela %>%
  mutate(epigenetic_measure = rename_clocks(rename_accel(rename_scores(epigenetic_measure)))) %>%
  mutate(epigenetic_measure = case_when(epigenetic_measure != "DunedinPACE" 
                                        & epigenetic_measure %in% clockorder ~ paste0(epigenetic_measure, "Accel"),
                                        TRUE ~ epigenetic_measure))

# sort & round values
epiorder <- c("HorvathAccel", "PCHorvathAccel", "SkinBloodAccel", "PCSkinBloodAccel", "HannumAccel", "PCHannumAccel", "PedBEAccel", "WuAccel", "PhenoAgeAccel", "PCPhenoAgeAccel", "GrimAgeAccel", "PCGrimAgeAccel", "GrimAge2Accel", "DunedinPACE", "BMI (Do, 2023)",  "BMI (Wahl, 2017)", "CRP (Wielscher, 2022)", "CRP (Ligthart, 2016)", "Epigenetic-g", "DNAmTL", "PC DNAmTL")
results_ela <- results_ela %>%
  arrange(factor(epigenetic_measure, levels = epiorder)) %>%
  round_values()
```

## Socioeconomic status (SES) effects 

### Fit linear models in blood & saliva with and without correction for estimated cell type proportions and smoking exposure

```{r}
# create empty results object
results_ses <- data.frame(matrix(nrow = 0, ncol = 9))
colnames(results_ses) <- c("epigenetic_measure", 
                           "beta_std_blood_minimal_model", "p_blood_minimal_model", 
                           "beta_std_blood_smoking_celltype_adjusted", "p_blood_smoking_celltype_adjusted", 
                           "beta_std_saliva_minimal_model", "p_saliva_minimal_model",
                           "beta_std_saliva_smoking_celltype_adjusted", "p_saliva_smoking_celltype_adjusted")

# loop linear model over all epigenetic measures in blood and saliva
for(epimeasure in epimeasures){
  # subset blood data
  current_data_blood = select(all_cross_tissue_inner, 
                              paste0(epimeasure, "_blood"), child_id, ses_total_negative,
                              bio_blood_age, sex, 
                              cg05575921_blood,
                              CD4Tnv_blood, Baso_blood, CD4Tmem_blood, Bmem_blood, 
                              Bnv_blood, Treg_blood, CD8Tmem_blood, CD8Tnv_blood, 
                              Eos_blood, NK_blood, Mono_blood) %>%
    dplyr::rename("epimeasure" = paste0(epimeasure, "_blood")) %>%
    # standardize all numerical variables to mean = 0, sd = 1 (to get standardized betas)
    mutate(across(where(is.numeric), scale))
  # fit basic model in blood (only adjusted for age, sex, kinship)
  fit_blood_minimal = lmekin(
    formula = epimeasure ~ ses_total_negative + bio_blood_age + sex +
         (1 | child_id),
       data = current_data_blood,
       na.action = "na.omit", 
       varlist = kinship)
  sumstats_blood_minimal = extract_coxme_table(fit_blood_minimal)
  # extract results blood minimal model
  coefs = data.frame(matrix(nrow = 1, ncol = 0))
  coefs$epigenetic_measure = epimeasure
  coefs$beta_std_blood_minimal_model = sumstats_blood_minimal["ses_total_negative",]$beta
  coefs$p_blood_minimal_model = sumstats_blood_minimal["ses_total_negative",]$p
  # fit model in blood adjusted for age, sex, kinship, epigenetic smoke exposure proxy, estimated cell types
  fit_blood_full = lmekin(
    formula = epimeasure ~ ses_total_negative + bio_blood_age + sex +
      cg05575921_blood +
    CD4Tnv_blood + Baso_blood + CD4Tmem_blood + Bmem_blood + Bnv_blood + Treg_blood + 
    CD8Tmem_blood + CD8Tnv_blood + Eos_blood + NK_blood + Mono_blood +
         (1 | child_id),
       data = current_data_blood,
       na.action = "na.omit", 
       varlist = kinship)
  sumstats_blood_full = extract_coxme_table(fit_blood_full)
  # extract results blood full model
  coefs$epigenetic_measure = epimeasure
  coefs$beta_std_blood_smoking_celltype_adjusted = sumstats_blood_full["ses_total_negative",]$beta
  coefs$p_blood_smoking_celltype_adjusted = sumstats_blood_full["ses_total_negative",]$p
  # subset saliva data
  current_data_saliva = select(all_cross_tissue_inner, 
                        paste0(epimeasure, "_saliva"), child_id, ses_total_negative,
                        bio_saliva_age, sex, 
                        cg05575921_saliva, smoking_exposure_score_saliva,
                        Epithelial_saliva) %>%
    dplyr::rename("epimeasure" = paste0(epimeasure, "_saliva")) %>%
    # standardize all numerical variables to mean = 0, sd = 1 (to get standardized betas)
    mutate(across(where(is.numeric), scale))
  # fit basic model in saliva (only adjusted for age, sex, kinship)
  fit_saliva_minimal = lmekin(
    formula = epimeasure ~ ses_total_negative + bio_saliva_age + sex +
         (1 | child_id),
       data = current_data_saliva,
       na.action = "na.omit", 
       varlist = kinship)
  sumstats_saliva_minimal = extract_coxme_table(fit_saliva_minimal)
  # extract results saliva minimal model
  coefs$epigenetic_measure = epimeasure
  coefs$beta_std_saliva_minimal_model = sumstats_saliva_minimal["ses_total_negative",]$beta
  coefs$p_saliva_minimal_model = sumstats_saliva_minimal["ses_total_negative",]$p
  # fit model in saliva adjusted for age, sex, kinship, epigenetic smoke exposure proxy, estimated cell types
  fit_saliva_full = lmekin(
    formula = epimeasure ~ ses_total_negative + bio_saliva_age + sex +
      cg05575921_saliva +
      Epithelial_saliva +
         (1 | child_id),
       data = current_data_saliva,
       na.action = "na.omit", 
       varlist = kinship)
  sumstats_saliva_full = extract_coxme_table(fit_saliva_full)
  # extract results saliva full model
  coefs$epigenetic_measure = epimeasure
  coefs$beta_std_saliva_smoking_celltype_adjusted = sumstats_saliva_full["ses_total_negative",]$beta
  coefs$p_saliva_smoking_celltype_adjusted = sumstats_saliva_full["ses_total_negative",]$p
  # append results
  results_ses <- rbind(results_ses, coefs)
}

# FDR-correct results
fdr_ses <- as.data.frame(matrix(nrow = 2*nrow(results_ses), ncol = 0))
fdr_ses$tissue <- c(rep("blood", nrow(results_ses)), 
                   rep("saliva", nrow(results_ses)))
fdr_ses$p_fdr_minimal_model <- p.adjust(c(results_ses$p_blood_minimal_model, 
                                          results_ses$p_saliva_minimal_model), method = "fdr")
fdr_ses$p_fdr_smoking_celltype_adjusted <- p.adjust(c(results_ses$p_blood_smoking_celltype_adjusted,
                                                      results_ses$p_saliva_smoking_celltype_adjusted), method = "fdr")
# append FDR-corrected results
results_ses <- results_ses %>%
  mutate(p_fdr_blood_minimal_model = filter(fdr_ses, tissue == "blood")$p_fdr_minimal_model,
         p_fdr_saliva_minimal_model = filter(fdr_ses, tissue == "saliva")$p_fdr_minimal_model,
         p_fdr_blood_smoking_celltype_adjusted = filter(fdr_ses, tissue == "blood")$p_fdr_smoking_celltype_adjusted,
         p_fdr_saliva_smoking_celltype_adjusted = filter(fdr_ses, tissue == "saliva")$p_fdr_smoking_celltype_adjusted) %>%
  relocate(p_fdr_blood_minimal_model, .after = "p_blood_minimal_model") %>%
  relocate(p_fdr_blood_smoking_celltype_adjusted, .after = "p_blood_smoking_celltype_adjusted")

# rename epigenetic clocks/scores
results_ses <- results_ses %>%
  mutate(epigenetic_measure = rename_clocks(rename_accel(rename_scores(epigenetic_measure)))) %>%
  mutate(epigenetic_measure = case_when(epigenetic_measure != "DunedinPACE" 
                                        & epigenetic_measure %in% clockorder ~ paste0(epigenetic_measure, "Accel"),
                                        TRUE ~ epigenetic_measure))

# sort & round values
epiorder <- c("HorvathAccel", "PCHorvathAccel", "SkinBloodAccel", "PCSkinBloodAccel", "HannumAccel", "PCHannumAccel", "PedBEAccel", "WuAccel", "PhenoAgeAccel", "PCPhenoAgeAccel", "GrimAgeAccel", "PCGrimAgeAccel", "GrimAge2Accel", "DunedinPACE", "BMI (Do, 2023)",  "BMI (Wahl, 2017)", "CRP (Wielscher, 2022)", "CRP (Ligthart, 2016)", "Epigenetic-g", "DNAmTL", "PC DNAmTL")
results_ses <- results_ses %>%
  arrange(factor(epigenetic_measure, levels = epiorder)) %>%
  round_values()
```

## Export all results and transform to long format for plotting
```{r}
# combine & export results
results_ela <- results_ela %>%
  mutate(exposure = rep("Early-life adversity", nrow(results_ela))) %>%
  relocate(exposure, .after = epigenetic_measure)
results_ses <- results_ses %>%
  mutate(exposure = rep("Lower SES", nrow(results_ses))) %>%
  relocate(exposure, .after = epigenetic_measure)
results_exposures <- rbind(results_ela, results_ses)
save(results_exposures, file = "./03_results/results_exposures.Rdata")
write_xlsx(results_exposures, path = "05_tables_for_publication/supptable_10_exposure_associations.xlsx", col_names = TRUE, format_headers = TRUE)

## results to long format & add significance category
# celltype-adjusted
results_exposures_long_celltype <- results_exposures %>%
    select(epigenetic_measure, exposure, ends_with("_smoking_celltype_adjusted"))
colnames(results_exposures_long_celltype) <- str_remove_all(colnames(results_exposures_long_celltype), "_smoking_celltype_adjusted")
results_exposures_long_celltype <- results_exposures_long_celltype %>%
    pivot_longer(
    cols = ends_with("_blood") | ends_with("_saliva"),
    names_to = c(".value", "tissue"),
    names_pattern = "(.*)_(blood|saliva)"
  ) %>%
  mutate(significance = case_when(
    p < 0.05 & p_fdr < 0.05 ~ "significant",
    p < 0.05 & p_fdr >= 0.05 ~ "nominal",
    TRUE ~ "n.s.")) %>%
  mutate(effect = paste0(tissue, ": ", significance),
         type = case_when(epigenetic_measure %in% c("HorvathAccel", "PCHorvathAccel",
                                                    "SkinBloodAccel", "PCSkinBloodAccel",
                                                    "HannumAccel", "PCHannumAccel", 
                                                    "WuAccel", "PedBEAccel") 
                          ~ "Chronological clocks",
                          epigenetic_measure %in% c("PhenoAgeAccel", "PCPhenoAgeAccel",
                                                     "GrimAgeAccel", "PCGrimAgeAccel",
                                                     "GrimAge2Accel", "DunedinPACE") 
                           ~ "Biological clocks",
                           epigenetic_measure %in% c("BMI (Do, 2023)", 
                                                     "BMI (Wahl, 2017)",
                                                     "CRP (Wielscher, 2022)",
                                                     "CRP (Ligthart, 2016)", 
                                                     "Epigenetic-g", "DNAmTL",
                                                     "PC DNAmTL")
                           ~ "Epigenetic scores"))

results_exposures_long_minimal <- results_exposures %>%
    select(epigenetic_measure, exposure, ends_with("_minimal_model"))
colnames(results_exposures_long_minimal) <- str_remove_all(colnames(results_exposures_long_minimal), "_minimal_model")
results_exposures_long_minimal <- results_exposures_long_minimal %>%
    pivot_longer(
    cols = ends_with("_blood") | ends_with("_saliva"),
    names_to = c(".value", "tissue"),
    names_pattern = "(.*)_(blood|saliva)"
  ) %>%
  mutate(significance = case_when(
    p < 0.05 & p_fdr < 0.05 ~ "significant",
    p < 0.05 & p_fdr >= 0.05 ~ "nominal",
    TRUE ~ "n.s.")) %>%
  mutate(effect = paste0(tissue, ": ", significance),
         type = case_when(epigenetic_measure %in% c("HorvathAccel", "PCHorvathAccel",
                                                    "SkinBloodAccel", "PCSkinBloodAccel",
                                                    "HannumAccel", "PCHannumAccel", 
                                                    "WuAccel", "PedBEAccel") 
                          ~ "Chronological clocks",
                          epigenetic_measure %in% c("PhenoAgeAccel", "PCPhenoAgeAccel",
                                                     "GrimAgeAccel", "PCGrimAgeAccel",
                                                     "GrimAge2Accel", "DunedinPACE") 
                           ~ "Biological clocks",
                           epigenetic_measure %in% c("BMI (Do, 2023)", 
                                                     "BMI (Wahl, 2017)",
                                                     "CRP (Wielscher, 2022)",
                                                     "CRP (Ligthart, 2016)", 
                                                     "Epigenetic-g", "DNAmTL",
                                                     "PC DNAmTL")
                           ~ "Epigenetic scores"))

```

# Plot ELA and SES effects with full correction (Figure 6)

```{r}
# horizontal barplot standardized betas by measure, color = tissue, bar alpha = significance (n.s., nominal, FDR), facets = exposure
results_barplot <- ggplot(aes(x = beta_std, 
                          y = factor(epigenetic_measure, levels = rev(epiorder)), 
                          color = factor(tissue, levels = c("saliva", "blood")), 
                          fill = factor(tissue, levels = c("saliva", "blood")),
                          pattern_density = significance),
                      data = results_exposures_long_celltype) +
  # add hlines to separate groups of epigenetic measures
  geom_hline(yintercept = c(7.5, 13.5), color = "grey", linewidth = 0.5) +
  # add hlines to separate effects between clocks
  geom_hline(yintercept = seq(1.5,21.5, 1), color = "lightgrey", linewidth = 0.25) +
  # barplot
  geom_bar_pattern(stat = "identity", position="dodge2", 
                   pattern = "stripe", pattern_fill = "white", pattern_color = "white",
                   pattern_key_scale_factor=.5, linewidth = 0.9, width = 0.8) +
  ylab("Epigenetic scores      Biological clocks      Chronological clocks") +
  xlab("Effect size (standardized beta)") +
  scale_color_brewer("Tissue", palette = "Set1", direction = -1, guide = "none") +
  scale_fill_brewer("Tissue", palette = "Set1", direction = -1, 
                    guide = guide_legend(reverse = TRUE, override.aes = list(pattern_density = 0))) +
  scale_pattern_density_manual(values = c("n.s." = 1, "nominal" = 0.3, "significant" = 0), guide = guide_legend(override.aes = list(color = "grey"))) +
  labs(pattern_density = "Significance") +
  # scale_fill_manual(name = "Effect", values = c("#fce8e8", "#f28c8e", "#E41A1C", "#ebf2f8", "#9bbedc", "#377EB8")) +
  theme_bw() +
  theme(panel.grid.major.y = element_blank()) +
  facet_wrap(facets = "exposure", nrow = 1, ncol = 2) +
  theme(strip.text.x = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        axis.title.y =  element_text(size = 12, face = "bold"),
    strip.background = element_rect(fill="white", color = "white"),
    legend.position = "inside", legend.position.inside = c(0.6, 0.81))

results_barplot

png(file = "04_figures/figure_6_combined_barplot_exposures_smoking_celltype_adjusted.png", height = 7, width = 10, units = 'in', res = 700)
results_barplot
dev.off()
```

# Plot ELA and SES effects without cell type and smoking correction (Supplementary Figure 4)

```{r}
# horizontal barplot standardized betas by measure, color = tissue, bar alpha = significance (n.s., nominal, FDR), facets = exposure
results_barplot <- ggplot(aes(x = beta_std, 
                          y = factor(epigenetic_measure, levels = rev(epiorder)), 
                          color = factor(tissue, levels = c("saliva", "blood")), 
                          fill = factor(tissue, levels = c("saliva", "blood")), 
                          pattern_density = significance),
                      data = results_exposures_long_minimal) +
  # add hlines to separate groups of epigenetic measures
  geom_hline(yintercept = c(7.5, 13.5), color = "grey", linewidth = 0.5) +
  # add hlines to separate effects between clocks
  geom_hline(yintercept = seq(1.5,21.5, 1), color = "lightgrey", linewidth = 0.25) +
  # barplot
  geom_bar_pattern(stat = "identity", position="dodge2", 
                   pattern = "stripe", pattern_fill = "white", pattern_color = "white",
                   pattern_key_scale_factor=.5, linewidth = 0.9, width = 0.8) +
  ylab("Epigenetic scores      Biological clocks      Chronological clocks") +
  xlab("Effect size (standardized beta)") +
  ggtitle("Minimal model") +
  scale_color_brewer("Tissue", palette = "Set1", direction = -1, guide = "none") +
  scale_fill_brewer("Tissue", palette = "Set1", direction = -1, 
                    guide = guide_legend(reverse = TRUE, override.aes = list(pattern_density = 0))) +
  scale_pattern_density_manual(values = c("n.s." = 1, "nominal" = 0.3, "significant" = 0), guide = guide_legend(override.aes = list(color = "grey"))) +
  labs(pattern_density = "Significance") +
  theme_bw() +
  theme(panel.grid.major.y = element_blank()) +
  facet_wrap(facets = "exposure", nrow = 1, ncol = 2) +
  theme(strip.text.x = element_text(size = 12, face = "bold"),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        axis.title.y =  element_text(size = 12, face = "bold"),
    strip.background = element_rect(fill="white", color = "white"),
    legend.position = "inside", legend.position.inside = c(0.6, 0.81))

results_barplot

png(file = "04_figures/supplementary_figure_4_combined_barplot_exposures_minimal_model.png", height = 7, width = 10, units = 'in', res = 700)
results_barplot
dev.off()
```
