---
title: 'Evaluating epigenetic clocks and scores: Biomarker associations'
author: "Vera N. Karlbauer"
contact: "vera_karlbauer@psych.mpg.de"
date: "2026-08-09"
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/vera_karlbauer/Documents/Kids2Health/02_stats/02_analysis/evaluating_epigenetic_clocks_and_scores", digits = 4, scipen = 6)
```

# Purpose:  
- As additional benchmarks for epigenetic clocks, test associations with clinical cardiometabolic markers.
- Compare effect sizes between tissues.
- Epigenetic measures: 14 epigenetic clocks, chronological & biological.
- Clinical markers: Systolic blood pressure, BodPod fat mass, metabolites from Nightingale CoreMetabolomics panel (based on fasting plasma):
  - Total cholesterol
  - HDL cholesterol 
  - LDL cholesterol
  - Total triglycerides
- Covariates: age, sex, kinship matrix (ancestry & relatedness).
- Additional control for estimated cell type composition (analyses run with and without). 
- FDR-correct over all clocks per biomarker.
- Plot results as barplot (Figure 7 and Supplementary Figure 5).
- Export results as Supplementary Table 11.

# Setup
### Step 0: Load packages

```{r}
library(rmarkdown)
library(dplyr)
library(RColorBrewer)
library(coxme)
library(writexl)
library(tidyr)
library(ggplot2)
library(stringr)
library(ggpubr)
library(ggpattern)
```


### Step 1: Source functions and utilities

```{r}
source("./00_functions/01_functions.R")
source("./00_functions/02_utilities.R")
```

### Step 2: Load data

```{r, include = FALSE}
# main data
load("../../../01_data/02_proc/09_cross_tissue/all_cross_tissue_inner.Rdata")
# kinship matrix
load("../../../01_data/02_proc/02_genotypes/qc-imputation_output/genotype_data/01_qc/13_ancestry_rels/K2H_SP1_2_3_4_genetic_kinship_matrix.Rdata")
```

### Step 3: Prepare data 

```{r}
# filter data
all_cross_tissue_inner <- all_cross_tissue_inner %>%
  filter(child_id %in% rownames(k2H_pcrelate_kinship_matrix)) %>%
  filter(timepoint == "T0")

# filter kinship matrix
kinship <- k2H_pcrelate_kinship_matrix[unique(all_cross_tissue_inner$child_id), unique(all_cross_tissue_inner$child_id)]
```

# Analysis: Clinical biomarker associations

### Loop associations across all clocks and biomarkers in blood and saliva with and without correction for smoke exposure and estimated cell type composition

```{r}
# define epigenetic measures
epimeasures <- gsub("_saliva", "", 
                    colnames(all_cross_tissue_inner %>%
                               select(contains("saliva")) %>%
                               select(starts_with("epiage_accel") 
                                      | starts_with("dunedin_pace")) %>%
                               select(!contains("resid"))))
# define biomarkers
biomarkers <- c("bp_systolic", "bodpod_fatmass", "TotalC", "LDLC", "HDLC", "TotalTG")
# create empty results object
results_biomarkers <- data.frame(matrix(nrow = 0, ncol = 11))
colnames(results_biomarkers) <- c("epigenetic_measure", "biomarker", "sample_size",
                                  "beta_std_blood_minimal_model", "p_blood_minimal_model",
                                  "beta_std_blood_smoking_celltype_adjusted",
                                  "p_blood_smoking_celltype_adjusted", 
                                  "beta_std_saliva_minimal_model", "p_saliva_minimal_model",
                                  "beta_std_saliva_smoking_celltype_adjusted",
                                  "p_saliva_smoking_celltype_adjusted")

# loop linear model over all epigenetic measures and biomarkers in blood and saliva
for(marker in biomarkers){
  for(epimeasure in epimeasures){
    # subset blood data
    current_data_blood = select(all_cross_tissue_inner, 
                              paste0(epimeasure, "_blood"), marker,
                              child_id,
                              bio_blood_age, sex, 
                              cg05575921_blood,
                              CD4Tnv_blood, Baso_blood, CD4Tmem_blood, Bmem_blood,
                              Bnv_blood, Treg_blood, CD8Tmem_blood, CD8Tnv_blood,
                              Eos_blood, NK_blood, Mono_blood) %>%
      dplyr::rename("epimeasure" = paste0(epimeasure, "_blood")) %>%
      dplyr::rename("marker" = marker) %>%
      # standardize all numerical variables to mean = 0, sd = 1 (to get standardized betas)
      mutate(across(where(is.numeric), scale))
    # fit basic model in blood (only adjusted for age, sex, kinship)
    fit_blood_minimal = lmekin(formula = marker ~ epimeasure + bio_blood_age + sex + 
                         (1 | child_id),
                       data = current_data_blood,
                       na.action = "na.omit", 
                       varlist = kinship)
    sumstats_blood_minimal = extract_coxme_table(fit_blood_minimal)
    # extract results blood minimal
    coefs = data.frame(matrix(nrow = 1, ncol = 0))
    coefs$epigenetic_measure = epimeasure
    coefs$biomarker = marker
    coefs$sample_size = nrow(filter(current_data_blood, !is.na(marker)))
    coefs$beta_std_blood_minimal_model = sumstats_blood_minimal["epimeasure",]$beta
    coefs$p_blood_minimal_model = sumstats_blood_minimal["epimeasure",]$p
    # fit model in blood adjusted for age, sex, kinship, epigenetic smoke exposure proxy, estimated cell types
    fit_blood_full = lmekin(formula = marker ~ epimeasure + bio_blood_age + sex + 
                         cg05575921_blood +
                         CD4Tnv_blood + Baso_blood + CD4Tmem_blood + Bmem_blood + 
                         Bnv_blood + Treg_blood + CD8Tmem_blood + CD8Tnv_blood + 
                         Eos_blood + NK_blood + Mono_blood +
                         (1 | child_id),
                       data = current_data_blood,
                       na.action = "na.omit", 
                       varlist = kinship)
    sumstats_blood_full = extract_coxme_table(fit_blood_full)
    # extract results blood full
    coefs$beta_std_blood_smoking_celltype_adjusted = sumstats_blood_full["epimeasure",]$beta
    coefs$p_blood_smoking_celltype_adjusted = sumstats_blood_full["epimeasure",]$p
    # subset saliva data
    current_data_saliva = select(all_cross_tissue_inner, 
                              paste0(epimeasure, "_saliva"), marker,
                              child_id,
                              bio_saliva_age, sex,
                              cg05575921_saliva,
                              Epithelial_saliva) %>%
      dplyr::rename("epimeasure" = paste0(epimeasure, "_saliva")) %>%
      dplyr::rename("marker" = marker) %>%
      # standardize all numerical variables to mean = 0, sd = 1 (to get standardized betas)
      mutate(across(where(is.numeric), scale))
    # fit basic model in saliva (only adjusted for age, sex, kinship)
    fit_saliva_minimal = lmekin(formula = marker ~ epimeasure + bio_saliva_age + sex + 
                         (1 | child_id),
                       data = current_data_saliva,
                       na.action = "na.omit", 
                       varlist = kinship)
    sumstats_saliva_minimal = extract_coxme_table(fit_saliva_minimal)
    # extract results saliva minimal
    coefs$beta_std_saliva_minimal_model = sumstats_saliva_minimal["epimeasure",]$beta
    coefs$p_saliva_minimal_model = sumstats_saliva_minimal["epimeasure",]$p
    # fit full model in saliva adjusted for age, sex, kinship, epigenetic smoke exposure proxy, estimated cell types
    fit_saliva_full = lmekin(formula = marker ~ epimeasure + bio_saliva_age + sex + 
                          cg05575921_saliva + Epithelial_saliva + 
                          (1 | child_id),
                       data = current_data_saliva,
                       na.action = "na.omit", 
                       varlist = kinship)
    sumstats_saliva_full = extract_coxme_table(fit_saliva_full)
    # extract results saliva
    coefs$beta_std_saliva_smoking_celltype_adjusted  = sumstats_saliva_full["epimeasure",]$beta
    coefs$p_saliva_smoking_celltype_adjusted  = sumstats_saliva_full["epimeasure",]$p
    # append results
    results_biomarkers <- rbind(results_biomarkers, coefs)
  }
}
# FDR-correct per biomarker across all clocks & tissues
pvals_all <- data.frame(matrix(nrow = 1, ncol = 0))
for(marker in biomarkers){
    pvals = results_biomarkers %>% 
    filter(biomarker == marker) %>% 
    select(epigenetic_measure, biomarker, 
           p_blood_minimal_model, p_blood_smoking_celltype_adjusted,
           p_saliva_minimal_model, p_saliva_smoking_celltype_adjusted) %>%
    pivot_longer(cols = starts_with("p_blood") | starts_with("p_saliva"),
                 names_to = c("tissue", ".value"),
                 names_pattern = "p_(blood|saliva)_(.+)")
    pvals$p_fdr_minimal_model = p.adjust(pvals$minimal_model, method = "fdr")
    pvals$p_fdr_smoking_celltype_adjusted = p.adjust(pvals$smoking_celltype_adjusted, method = "fdr")
    pvals = pvals %>%
      select(epigenetic_measure, biomarker, tissue, starts_with("p_fdr")) %>%
      pivot_wider(values_from = starts_with("p_fdr"),
                  names_from = "tissue",
                  names_glue = "p_fdr_{tissue}_{sub('p_fdr_', '', .value)}")
    # append corrected p-values
    pvals_all <- rbind(pvals_all, pvals)
}
results_biomarkers <- left_join(results_biomarkers, pvals_all, by = c("epigenetic_measure", "biomarker"))

# rename epigenetic clocks/scores
results_biomarkers <- results_biomarkers %>%
  mutate(epigenetic_measure = rename_clocks(rename_accel(epigenetic_measure))) %>%
  mutate(epigenetic_measure = case_when(epigenetic_measure != "DunedinPACE" 
                                        & epigenetic_measure %in% clockorder ~
                                          paste0(epigenetic_measure, "Accel"),
                                        TRUE ~ epigenetic_measure))
# arrange
epiorder <- c("HorvathAccel", "PCHorvathAccel", "SkinBloodAccel", "PCSkinBloodAccel", "HannumAccel", "PCHannumAccel", "PedBEAccel", "WuAccel", "PhenoAgeAccel", "PCPhenoAgeAccel", "GrimAgeAccel", "PCGrimAgeAccel", "GrimAge2Accel", "DunedinPACE")
results_biomarkers <- results_biomarkers %>%
    arrange(factor(epigenetic_measure, levels = epiorder)) %>%
  relocate(p_fdr_blood_minimal_model, .after = p_blood_minimal_model) %>%
  relocate(p_fdr_saliva_minimal_model, .after = p_saliva_minimal_model) %>%
  relocate(p_fdr_blood_smoking_celltype_adjusted, .after = p_blood_smoking_celltype_adjusted) %>%
  relocate(p_fdr_saliva_smoking_celltype_adjusted, .after = p_saliva_smoking_celltype_adjusted)
# round values
results_biomarkers <- round_values(results_biomarkers)
# rename biomarkers
results_biomarkers <- results_biomarkers %>%
  mutate(biomarker = case_match(biomarker,
                                "bp_systolic" ~ "Systolic blood pressure",
                                "bodpod_fatmass" ~ "Body fat mass",
                                "TotalC" ~ "Total cholesterol",
                                "LDLC" ~ "LDL cholesterol",
                                "HDLC" ~ "HDL cholesterol",
                                "TotalTG" ~ "Total triglycerides",
                                 .default = biomarker))
```

### Export results as Supplementary Table 11
```{r}
save(results_biomarkers, file = "./03_results/results_biomarkers.Rdata")
write_xlsx(results_biomarkers, path = "05_tables_for_publication/supptable_11_biomarker_associations.xlsx", col_names = TRUE, format_headers = TRUE)
```

## Plot results

### Without cell type correction (minimal model) - Supplementary Figure 5
```{r}
# results to long format & add significance category
results_biomarkers_long <- results_biomarkers %>%
      select(epigenetic_measure, biomarker, ends_with("_minimal_model"))
colnames(results_biomarkers_long) <- str_remove_all(colnames(results_biomarkers_long), "_minimal_model")
results_biomarkers_long <- results_biomarkers_long %>%
  pivot_longer(
    cols = contains("_blood") | contains("_saliva"),
    names_to = c(".value", "tissue"),
    names_pattern = "^(.*)_(blood|saliva)$") %>%
  mutate(significance = case_when(
    p < 0.05 & p_fdr < 0.05 ~ "significant",
    p < 0.05 & p_fdr >= 0.05 ~ "nominal",
    p >= 0.05 ~ "n.s.",
    TRUE ~ "n.s.")) %>%
  mutate(effect = paste0(tissue, ": ", significance),
         type = case_when(epigenetic_measure %in% c("HorvathAccel", "PCHorvathAccel",
                                                    "SkinBloodAccel", "PCSkinBloodAccel",
                                                    "HannumAccel", "PCHannumAccel", 
                                                    "WuAccel", "PedBEAccel") 
                          ~ "Chronological clocks",
                          epigenetic_measure %in% c("PhenoAgeAccel", "PCPhenoAgeAccel",
                                                     "GrimAgeAccel", "PCGrimAgeAccel",
                                                     "GrimAge2Accel", "DunedinPACE") 
                           ~ "Biological clocks")) %>%
  arrange(factor(biomarker, levels = c("Systolic blood pressure",
                                       "Body fat mass",
                                       "Total cholesterol",
                                       "LDL cholesterol",
                                       "HDL cholesterol",
                                       "Total triglycerides")))
# y axis annotation
axis_annotation_df <- data.frame(
  biomarker = factor(
    rep(c("Systolic blood pressure", "Total cholesterol", "HDL cholesterol"), each = 2),
    levels = c("Systolic blood pressure", "Body fat mass",
               "Total cholesterol", "LDL cholesterol",
               "HDL cholesterol", "Total triglycerides")),
  y = rep(c(10, 3.5), 3),
  label = rep(c("Chronological clocks", "Biological clocks"), 3))

# horizontal barplot standardized betas by measure, color = tissue, bar alpha = significance (n.s., nominal, FDR), facets = biomarker
biomarker_barplot <- ggplot(aes(x = beta_std, 
                          y = factor(epigenetic_measure, levels = rev(epiorder)), 
                          color = factor(tissue, levels = c("saliva", "blood")), 
                          fill = factor(tissue, levels = c("saliva", "blood")),
                          pattern_density = significance),
                      data = results_biomarkers_long) +
  # add hline to separate groups of epigenetic measures
  geom_hline(yintercept = 6.5, color = "grey", linewidth = 0.5) +
  # add hlines to separate effects between clocks
  geom_hline(yintercept = seq(1.5, 13.5, 1), color = "lightgrey", linewidth = 0.25) +
  # barplot
  # geom_bar(stat = "identity", position="dodge2") +
  geom_bar_pattern(stat = "identity", position="dodge2", 
                   pattern = "stripe", pattern_fill = "white", pattern_color = "white",
                   pattern_key_scale_factor=.5, linewidth = 0.9, width = 0.8) +
  xlab("Effect size (standardized beta)") +
  ggtitle("Minimal model") +
  scale_color_brewer("Tissue", palette = "Set1", direction = -1, guide = "none") +
  scale_fill_brewer("Tissue", palette = "Set1", direction = -1, 
                    guide = guide_legend(reverse = TRUE, 
                                         override.aes = list(pattern_density = 0))) +
  scale_pattern_density_manual("Significance", values = c("n.s." = 1, "nominal" = 0.3, "significant" = 0), guide = guide_legend(override.aes = list(color = "grey"))) +
  theme(panel.grid.major.y = element_blank()) +
  facet_wrap(~factor(biomarker, 
                     levels = c("Systolic blood pressure", "Body fat mass", 
                                "Total cholesterol","LDL cholesterol", 
                                "HDL cholesterol","Total triglycerides")),
             nrow = 3, ncol = 2) +
  geom_text(data = axis_annotation_df,
            aes(x = -Inf, y = y, label = label),
            inherit.aes = FALSE,
            angle = 90,
            vjust = -14,
            size = 5, fontface = "bold", color = GyPr_palette[20],
            lineheight = 0.85) +
  coord_cartesian(clip = "off") +
  theme_bw() +
  theme(strip.text.x = element_text(size = 12, face = "bold"),
        strip.background = element_rect(fill="white", color = "white"),
        axis.text.y = element_text(size = 12, face = "bold"),
        axis.title.y = element_blank(),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
        legend.position = "bottom",
        plot.margin = margin(t = 5, r = 5, b = 5, l = 55)) 

biomarker_barplot

png(file = "04_figures/supplementary_figure_5_barplot_biomarkers_minimal_model.png", height = 20, width = 14, units = 'in', res = 700)
biomarker_barplot
dev.off()
```

### With celltype and smoking correction - Figure 7
```{r}
# results to long format & add significance category
results_biomarkers_long <- results_biomarkers %>%
      select(epigenetic_measure, biomarker, ends_with("_smoking_celltype_adjusted"))
colnames(results_biomarkers_long) <- str_remove_all(colnames(results_biomarkers_long), "_smoking_celltype_adjusted")
results_biomarkers_long <- results_biomarkers_long %>%
  pivot_longer(
    cols = contains("_blood") | contains("_saliva"),
    names_to = c(".value", "tissue"),
    names_pattern = "^(.*)_(blood|saliva)$") %>%
  mutate(significance = case_when(
    p < 0.05 & p_fdr < 0.05 ~ "significant",
    p < 0.05 & p_fdr >= 0.05 ~ "nominal",
    p >= 0.05 ~ "n.s.",
    TRUE ~ "n.s.")) %>%
  mutate(effect = paste0(tissue, ": ", significance),
         type = case_when(epigenetic_measure %in% c("HorvathAccel", "PCHorvathAccel",
                                                    "SkinBloodAccel", "PCSkinBloodAccel",
                                                    "HannumAccel", "PCHannumAccel", 
                                                    "WuAccel", "PedBEAccel") 
                          ~ "Chronological clocks",
                          epigenetic_measure %in% c("PhenoAgeAccel", "PCPhenoAgeAccel",
                                                     "GrimAgeAccel", "PCGrimAgeAccel",
                                                     "GrimAge2Accel", "DunedinPACE") 
                           ~ "Biological clocks")) %>%
  arrange(factor(biomarker, levels = c("Systolic blood pressure",
                                       "Body fat mass",
                                       "Total cholesterol",
                                       "LDL cholesterol",
                                       "HDL cholesterol",
                                       "Total triglycerides")))

# horizontal barplot standardized betas by measure, color = tissue, bar alpha = significance (n.s., nominal, FDR), facets = biomarker
biomarker_barplot <- ggplot(aes(x = beta_std, 
                          y = factor(epigenetic_measure, levels = rev(epiorder)), 
                          color = factor(tissue, levels = c("saliva", "blood")), 
                          fill = factor(tissue, levels = c("saliva", "blood")),
                          pattern_density = significance),
                      data = results_biomarkers_long) +
  # add hline to separate groups of epigenetic measures
  geom_hline(yintercept = 6.5, color = "grey", linewidth = 0.5) +
  # add hlines to separate effects between clocks
  geom_hline(yintercept = seq(1.5, 13.5, 1), color = "lightgrey", linewidth = 0.25) +
  # barplot
  # geom_bar(stat = "identity", position="dodge2") +
  geom_bar_pattern(stat = "identity", position="dodge2", 
                   pattern = "stripe", pattern_fill = "white", pattern_color = "white",
                   pattern_key_scale_factor=.5, linewidth = 0.9, width = 0.8) +
  xlab("Effect size (standardized beta)") +
  scale_color_brewer("Tissue", palette = "Set1", direction = -1, guide = "none") +
  scale_fill_brewer("Tissue", palette = "Set1", direction = -1, 
                    guide = guide_legend(reverse = TRUE, override.aes = list(pattern_density = 0))) +
  scale_pattern_density_manual("Significance", values = c("n.s." = 1, "nominal" = 0.3, "significant" = 0), guide = guide_legend(override.aes = list(color = "grey"))) +
  theme(panel.grid.major.y = element_blank()) +
  facet_wrap(~factor(biomarker, 
                     levels = c("Systolic blood pressure", "Body fat mass", 
                                "Total cholesterol","LDL cholesterol", 
                                "HDL cholesterol","Total triglycerides")),
             nrow = 3, ncol = 2) +
  geom_text(data = axis_annotation_df,
            aes(x = -Inf, y = y, label = label),
            inherit.aes = FALSE,
            angle = 90,
            vjust = -14,
            size = 5, fontface = "bold", color = GyPr_palette[20],
            lineheight = 0.85) +
  coord_cartesian(clip = "off") +
  theme_bw() +
  theme(strip.text.x = element_text(size = 12, face = "bold"),
    strip.background = element_rect(fill="white", color = "white"),
    axis.text.y = element_text(size = 12, face = "bold"),
    axis.title.y = element_blank(),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.position = "bottom",
    plot.margin = margin(t = 5, r = 5, b = 5, l = 55))

biomarker_barplot

png(file = "04_figures/figure_7_barplot_biomarkers_smoking_celltype_adjusted.png", height = 20, width = 14, units = 'in', res = 700)
biomarker_barplot
dev.off()
```

