---
title: 'Evaluating epigenetic clocks and scores: Effects of PC clock imputation'
author: "Vera N. Karlbauer"
date: "2026-07-08"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(
  root.dir = "/Users/vera_karlbauer/Documents/Kids2Health/02_stats/02_analysis/evaluating_epigenetic_clocks_and_scores",
  digits = 4, scipen = 6)
```

# Purpose:
- Plot and examine differences in absolute clock performance (median absolute error MAE) between PC clocks with and without imputation; compared to non-PC counterparts (Extended Data Figure 1).  
- The standard settings for PC clock computation (source code by authors, ) perform median imputation of missing CpGs based on an adult blood reference. This might not be appropriate for applications in children or in other tissues.
- Therefore, imputation was manually disabled, and results with and without imputation are compared.

# Setup
### Step 0: Load packages
```{r, include = FALSE}
# load packages
library(rmarkdown)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)
library(RColorBrewer)
library(stringr)
library(Metrics)
```

### Step 1: Source functions and utilities

```{r, include = FALSE}
source("./00_functions/01_functions.R")
source("./00_functions/02_utilities.R")
```

### Step 2: Load data

```{r, include = FALSE}
# old estimation with PC clock imputation
load("../../../01_data/02_proc/09_cross_tissue/99_archive/all_cross_tissue_inner.Rdata")
# change dataset name immediately
old_with_imputation <- all_cross_tissue_inner

# new estimation without PC clock imputation
load("../../../01_data/02_proc/09_cross_tissue/all_cross_tissue_inner.Rdata")
```

### Step 3: Prepare data

```{r}
### Unimputed (new)
# select clocks only & transform to long format
new_epiclocks <- all_cross_tissue_inner %>%
  select(child_id, timepoint, subproject, bio_blood_age, bio_saliva_age, chrono_age, 
         starts_with(c("epiage_horvath", "epiage_pc_horvath", "epiage_skinblood", "epiage_pc_skinblood",
                       "epiage_hannum", "epiage_pc_hannum", "epiage_phenoage", "epiage_pc_phenoage",
                       "epiage_grimage", "epiage_pc_grimage"))) %>%
  select(!starts_with(c("epiage_accel_", "epiage_delta_", "epiage_grimage2", "epiage_accel_grimage2"))) %>%
  select(!contains("resid"))
# pivot by tissue
new_epiclocks_long <- pivot_longer(new_epiclocks,
                               starts_with("epiage"),
                               names_to = c(".value", "tissue"),
                               names_pattern = "^(.*)_(blood|saliva)$")
# pivot by clock name
new_epiclocks_long <- pivot_longer(new_epiclocks_long,
                               cols = starts_with("epiage"),
                               names_to = "clock_name",
                               names_prefix = "epiage_",
                               values_to = "epiage")

### Imputed (old)
# select clocks only & transform to long format
old_epiclocks <- old_with_imputation %>%
  select(child_id, timepoint, subproject, bio_blood_age, bio_saliva_age, chrono_age, 
         starts_with(c("epiage_horvath", "epiage_pc_horvath", "epiage_skinblood", "epiage_pc_skinblood",
                       "epiage_hannum", "epiage_pc_hannum", "epiage_phenoage", "epiage_pc_phenoage",
                       "epiage_grimage", "epiage_pc_grimage"))) %>%
  select(!starts_with(c("epiage_accel_", "epiage_delta_", "epiage_grimage2", "epiage_accel_grimage2"))) %>%
  select(!contains("resid"))
# pivot by tissue
old_epiclocks_long <- pivot_longer(old_epiclocks,
                               starts_with("epiage"),
                               names_to = c(".value", "tissue"),
                               names_pattern = "^(.*)_(blood|saliva)$")
# pivot by clock name
old_epiclocks_long <- pivot_longer(old_epiclocks_long,
                               cols = starts_with("epiage"),
                               names_to = "clock_name",
                               names_prefix = "epiage_",
                               values_to = "epiage")
```

# Analysis

## Calculate median absolute error per tissue per clock for imputed and unimputed PC clocks
### Imputed (old)
```{r}
old_mae_clocks <- as.data.frame(matrix(ncol = 3, nrow = 0))
colnames(old_mae_clocks) <- c("mae", "clock_name", "tissue")
clocknames <- unique(old_epiclocks_long$clock_name)

for(i in 1:length(clocknames)){
  clockname = clocknames[i]
  blood = filter(old_epiclocks_long, tissue == "blood", clock_name == clockname)
  mae_blood = as.data.frame(mdae(actual = blood$bio_blood_age, predicted = blood$epiage))
  colnames(mae_blood) = c("mae")
  saliva = filter(old_epiclocks_long, tissue == "saliva", clock_name == clockname)
  mae_saliva = as.data.frame(mdae(actual = saliva$bio_saliva_age, predicted = saliva$epiage))
  colnames(mae_saliva) = c("mae")
  mae = rbind(mae_blood, mae_saliva)
  colnames(mae) = c("mae")
  mae$clock_name = c(clockname, clockname)
  mae$tissue = c("blood", "saliva")
  old_mae_clocks <- rbind(old_mae_clocks, mae)
}
old_mae_clocks <- old_mae_clocks %>%
 mutate(clock_name = str_replace_all(clock_name,
                                              c("pc_horvath" = "PCHorvath",
                                                "horvath" = "Horvath",
                                                "pc_skinblood" = "PCSkinBlood",
                                                "skinblood" = "SkinBlood",
                                                "pc_hannum" = "PCHannum",
                                                "hannum" = "Hannum",
                                                "pc_phenoage" = "PCPhenoAge",
                                                "phenoage" = "PhenoAge",
                                                "pc_grimage" = "PCGrimAge",
                                                 "grimage" = "GrimAge")))
colorder <- c("Horvath", "PCHorvath", "SkinBlood", "PCSkinBlood", "Hannum", "PCHannum", "PhenoAge", "PCPhenoAge", "GrimAge", "PCGrimAge")
old_mae_clocks <- old_mae_clocks[order(factor(old_mae_clocks$clock_name, levels = colorder)),]
```

### Unimputed (new)
```{r}
new_mae_clocks <- as.data.frame(matrix(ncol = 3, nrow = 0))
colnames(new_mae_clocks) <- c("mae", "clock_name", "tissue")
clocknames <- unique(new_epiclocks_long$clock_name)

for(i in 1:length(clocknames)){
  clockname = clocknames[i]
  blood = filter(new_epiclocks_long, tissue == "blood", clock_name == clockname)
  mae_blood = as.data.frame(mdae(actual = blood$bio_blood_age, predicted = blood$epiage))
  colnames(mae_blood) = c("mae")
  saliva = filter(new_epiclocks_long, tissue == "saliva", clock_name == clockname)
  mae_saliva = as.data.frame(mdae(actual = saliva$bio_saliva_age, predicted = saliva$epiage))
  colnames(mae_saliva) = c("mae")
  mae = rbind(mae_blood, mae_saliva)
  colnames(mae) = c("mae")
  mae$clock_name = c(clockname, clockname)
  mae$tissue = c("blood", "saliva")
  new_mae_clocks <- rbind(new_mae_clocks, mae)
}
new_mae_clocks <- new_mae_clocks %>%
 mutate(clock_name = str_replace_all(clock_name,
                                              c("pc_horvath" = "PCHorvath",
                                                "horvath" = "Horvath",
                                                "pc_skinblood" = "PCSkinBlood",
                                                "skinblood" = "SkinBlood",
                                                "pc_hannum" = "PCHannum",
                                                "hannum" = "Hannum",
                                                "pc_phenoage" = "PCPhenoAge",
                                                "phenoage" = "PhenoAge",
                                                "pc_grimage" = "PCGrimAge",
                                                 "grimage" = "GrimAge")))
new_mae_clocks <- new_mae_clocks[order(factor(new_mae_clocks$clock_name, levels = colorder)),]
```

## Plot MAE per clock per tissue for imputed and unimputed PC clocks (Extended Data Figure 1)
```{r}
# old: imputed
maeplot_old_imputed <- ggplot(old_mae_clocks, aes(x = clock_name, y = mae, fill = tissue)) +
  geom_bar(position = "dodge", stat="identity") +
  geom_text(aes(label=round(mae, digits = 2)), vjust=-0.4, color="black", size=3.5,
            position = position_dodge(0.9)) +
  scale_fill_manual(values=c("#EB5355", "#699ECA")) +
  scale_x_discrete(limit = colorder) +
  ylab("Median absolute error (years)") +
  xlab("Clock name") +
  ggtitle("MAE with imputation of missing CpGs on PC clocks") +
  # add abline to seperate chrono & bio clocks
  geom_vline(xintercept = 6.5, color = "darkgrey", linewidth = 0.5) +
  annotate("label", x=3, y=50, size = 4,
           label = "Chronological clocks",
           fill = "grey") +
  annotate("label", x=8.5, y=50, size = 4,
           label = "Biological clocks",
           fill = "grey") +
  ylim(0, 50) +
  theme_bw() +
  theme(legend.position = "inside", legend.position.inside = c(0.1,0.8),
        axis.text.x = element_text(angle = 45, vjust = 0.7, hjust = 0.7))
maeplot_old_imputed

# new: unimputed
maeplot_new_unimputed <- ggplot(new_mae_clocks, aes(x = clock_name, y = mae, fill = tissue)) +
  geom_bar(position = "dodge", stat="identity") +
  geom_text(aes(label=round(mae, digits = 2)), vjust=-0.4, color="black", size=3.5,
            position = position_dodge(0.9)) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_discrete(limit = colorder) +
  ylab("Median absolute error (years)") +
  xlab("Clock name") +
  ggtitle("MAE without imputation of missing CpGs on PC clocks") +
  # add abline to seperate chrono & bio clocks
  geom_vline(xintercept = 6.5, color = "darkgrey", linewidth = 0.5) +
  annotate("label", x=3, y=50, size = 4,
           label = "Chronological clocks",
           fill = "grey") +
  annotate("label", x=8.5, y=50, size = 4,
           label = "Biological clocks",
           fill = "grey") +
  ylim(0, 50) +
  theme_bw() +
  theme(legend.position = "inside", legend.position.inside = c(0.1,0.8),
        axis.text.x = element_text(angle = 45, vjust = 0.7, hjust = 0.7)) 
maeplot_new_unimputed

# create combined figure
combined = ggarrange(maeplot_old_imputed, maeplot_new_unimputed,
                        ncol = 1,
                        nrow = 2,
                        common.legend = FALSE,
  labels = c("A","B"))

png(filename = "./04_figures/supplementary_figure_1_pc_clock_imputation.png", width = 20, height = 28, res = 700, units = "cm")
  combined
dev.off()
```

## Examine MAE differences between PC clocks with and without imputation

### Calculate differences
```{r} 
maediff <- left_join(new_mae_clocks, old_mae_clocks, by = c("clock_name", "tissue"), suffix = c("_new_unimputed", "_old_imputed")) %>%
  relocate("mae_old_imputed", .after = "mae_new_unimputed") %>%
  mutate(maediff = mae_old_imputed - mae_new_unimputed)

maediff
```
